Predicting Fatalities in Toronto Car Crashes: Analyzing the Impact of Time, Vehicle Type, and Other Collision Factors

### 2024-12-03 paper/paper.qmd
added citations and updated intext referencing

The estimand of interest is the log-adjusted probability of a fatality occuring in a MVC. Specifically, we aim to quanitfy how specific environmental factors increase or decrease the likelihood of a fatality. By applying inferential analysis through Bayesian linear models, we assess not only the magnitude of these effects, but their underlying uncertainties (EDIT).

@@

The remainder of this paper is structured as follows: @sec-data describes the dataset and methodology and @sec-model exhibits the use of inferential models. @Sec-results presents the results of the analysis, detailing the observed relationships between likelihood of death and various circumstantial factors. @sec-discussion discusses the broader implications and limitations of our findings.@sec-appendix presents a detailed idealized methodology to improve data collection, and additional model summary and diagnostic information.

@@

## Overview

This dataset, "Police Annual Statistical Report - Traffic Collisions", was published and refreshed on October 21st, 2024, by the Toronto Police Service [insert citation]. The Toronto Police Service publishes various datasets on public safety and crime to inform the public about safety issues [@annual_statistics_report]. Data on traffic collisions is included in the Toronto Police Service's Annual Statistical Report, which also covers reported crimes, search of persons, firearms, and the Police Service's budget [@annual_statistics_report]. The data is collected using historical Motor Vehicle Collisions and classifies them into the following categories: 
* Property Damage (PD) Collisions
* Fail to Remain (FTR) Collisions, or commonly known as hit-and-run accidents
* Injury Collisions
* Fatalities

@@

There is an alternative dataset from the Toronto Police Service called "Motor Vehicle Collisions involving Killed or Seriously Injured Persons" (CITE). Unlike the alternative dataset, this paper's dataset focuses on all collisions, instead of only focusing on ones where someone was either killed or seriously injured. While the alternative dataset has more explanatory variables simply because more data is collected when someone dies or is seriously injured, this paper's aims to generalize if a fatality is more likely to occur based on the general circumstances surrounding a crash, such as the time of day or if property damage occurred. In addition, each entry in the alternative dataset represents a killed or injured person, meaning that there exists the possibility for duplicate entries. However, each entry in the selected dataset is a unique Motor Vehicle Collision, making it more suitable for the scope of this paper Thus, we ended up not going with the alternative dataset for this paper, but there are variables in the alternative dataset that may motivate future research on this subject. (EDIT TO MAKE MORE CLEAR)

The paper uses the R programming language [@citeR] to analyze the dataset. The tidyverse package was used to simulate the dataset. Also, the tidyverse [@citetidyverse], arrow [CITE] and opendatatoronto [@citeopendatatoronto] packages were used to download the Victims of Crime dataset. Then, the tidyverse [@citetidyverse] package was used to clean the raw dataset and generate tests. The testthat package [CITE THIS] was used to create tests for our cleaned dataset. Rstanarm [CITE], Arrow [CITE], and bayestestR [CITE] were used to create and test the model. Finally, ggplot2 [@citeggplot2], tidyverse [@citetidyverse], knitr [@citeknitr] and scales [@citescales] packages were used to create the tables and graphs to display the results. [edit this paragraph]

## Measurement

Transforming a real-life Motor Vehicle Collision to an entry in the dataset is a well-documented process by the Toronto Police Service. For insurance purposes, the Toronto Police Service requires drivers to fill out the Motor Vehicle Collision Report for any collisions that occur in Toronto (CITE). Drivers required to fill out a motor vehicle collision report if the combined damage is more than $2000, if someone is injured, if a criminal act such as a DUI occurs, or if a pedestrian is involved in the accident (CITE). These reports are retained for six years by the Toronto Police Service, with the exception of collisions resulting in a fatality, which are retained indefinitely. The form ensures documentation of collision characteristics, location, road condition, and the extent of damages, systematically recording the characteristics of each crash for further criminal investigation and data analysis.

@@

## Outcome variables

@@

```
However, after the transformation, the outcome variable now takes on binary values and the distribution and summary statistics are shown below in @fig-fatalites-cleaned and @tbl-fatalites:

@@

```
## Predictor variables

### Hour

@@

### Type of Crash

@@

### Automobile

@@

### Motorcycle

@@

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
### Passenger

@@

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
### Pedestrian

@@

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
### Police Division

@@

### Year

@@

### Interaction Terms

@@

#### Hour and Motorcycle

@@

#### Hour and Pedestrian

@@

# Display the summary statistics for pedestrians in a formatted table
summary_stats_pedestrian %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
# Model

## Model Set-Up and Justification

@@

## Model Assumptions and Limitations

@@

# Results

@@

# Discussion

@@

# Appendix {-}

## Exploration of Surveys, Sampling, and Observational Data

### Data Collection Methodology

The Toronto Police Service's Motor Vehicle Collision database is collected using a systematic process. When a collision occurs such that where persons are injured or a combined damage is valued at more than $\2000 to vehicles or property, or damage to any private, municipal or highway property, police officers complete a standardized Motor Vehicle Accident Report form. This data is then maintained by the City of Toronto, where the collection process follow a standardized process highlighted in the Motor Vehicle Collision Report manual (CITATION), ensuring reporting consistency. The dataset also defines a collision: the contact resulting from the motion of a motor vehicle or a street car of its load, which produces property damage, injury or death. A standardized definition ensures consistent reporting by the Toronto Police Service and highlights the dataset's clear boundaries.

For every accident that requires one to fill out the Motor Vehicle Accident Report form, the police officer fills out the SR-LD-402 form (CITATION). The process ensures consistency in reporting key details surrounding a motor vehicle accident.However, because police officers only need to fill out this form in specific instances, the data set only includes collisions where police officers were present or serious incidents that required a report. As a result, minor offences are usually left out in the dataset and often go unreported.

@@

### Idealized Methodology

@@

### Example Survey

@@

## Enhancement: Model Card

### Model Overview
This model predicts the occurrence of motor vehicle collisions (MVCs) based on various factors, including the time of day (hour), type of vehicle involved (motorcycle, pedestrian, automobile, etc.), and police division. The model is trained using data from 2015 to 2024, with a focus on the interactions between the hour of the day and motorcycle or pedestrian involvement.

### Model Details

@@

### Model Coefficients

@@

#### Interaction Terms

@@

### Model Performance Metrics

@@

### Training Time
- **Chain 1**:
  - Warm-up: 29.99 seconds
  - Sampling: 26.11 seconds
  - Total: 56.1 seconds
- **Chain 2**:
  - Warm-up: 19.62 seconds
  - Sampling: 17.23 seconds
  - Total: 36.84 seconds
- **Chain 3**:
  - Warm-up: 19.88 seconds
  - Sampling: 19.30 seconds
  - Total: 39.18 seconds
- **Chain 4**:
  - Warm-up: 22.40 seconds
  - Sampling: 19.17 seconds
  - Total: 41.57 seconds

### Conclusion
The model performs exceptionally well in predicting motor vehicle collisions based on hour, vehicle type, and other factors, with the highest predictive power for motorcycle-related accidents. Key interactions, such as between hour and motorcycle or pedestrian, highlight the importance of specific hours of the day in predicting collision occurrences. This model can be used for more effective traffic safety planning and intervention strategies.

@@

@fig-ppcheck compares observed data (dark line) with replicated posterior predictions (light blue lines). The close alignment suggests that the model accurately captures the data's central tendency and variability. @fig-convergencecheckTrace and @fig-convergencecheckRhat show that the sampling algorithm used, the Markov chain Monte Carlo (MCMC) algorithm, did not run into issues as the posterior distribution for the model was created. Using the checks presented by @citetellingstorieswithdata, both graphs do not show anything abnormal since the trace plots in @fig-convergencecheckTrace display substantial horizontal fluctuation across chains, indicating good mixing, while the Rhat values in @fig-convergencecheckRhat are close to 1 and well below 1.05, further supporting convergence.

>>>>

The estimand of interest is the log-adjusted probability of a fatality occurring in a MVC. Specifically, we aim to quantify how specific environmental factors increase or decrease the likelihood of a fatality. By applying inferential analysis through Bayesian linear models, we assess not only the magnitude of these effects, but their underlying uncertainties.

@@

The remainder of this paper is structured as follows: @sec-data describes the dataset and methodology and @sec-model exhibits the use of inferential models. @sec-results presents the results of the analysis, detailing the observed relationships between likelihood of death and various circumstantial factors. @sec-discussion discusses the broader implications and limitations of our findings.@sec-appendix presents a detailed idealized methodology to improve data collection, and additional model summary and diagnostic information.

@@

## Overview {#sec-data-overview}

This dataset, "Police Annual Statistical Report - Traffic Collisions", was published and refreshed on October 21st, 2024, by the Toronto Police Service [@vehiclecollision]. The Toronto Police Service publishes various datasets on public safety and crime to inform the public about safety issues [@annual_statistics_report]. Data on traffic collisions is included in the Toronto Police Service's Annual Statistical Report, which also covers reported crimes, search of persons, firearms, and the Police Service's budget [@annual_statistics_report]. The data is collected using historical Motor Vehicle Collisions and classifies them into the following categories: 
* Property Damage (PD) Collisions
* Fail to Remain (FTR) Collisions, or commonly known as hit-and-run accidents
* Injury Collisions
* Fatalities

@@

There is an alternative dataset from the Toronto Police Service called "Motor Vehicle Collisions involving Killed or Seriously Injured Persons" [@alternativedataset]. Unlike the alternative dataset, this paper's dataset focuses on all collisions, instead of only focusing on ones where someone was either killed or seriously injured. While the alternative dataset has more explanatory variables simply because more data is collected when someone dies or is seriously injured, this paper's aims to generalize if a fatality is more likely to occur based on the general circumstances surrounding a crash, such as the time of day or if property damage occurred. In addition, each entry in the alternative dataset represents a killed or injured person, meaning that there exists the possibility for duplicate entries. However, each entry in the selected dataset is a unique Motor Vehicle Collision, making it more suitable for the scope of this paper Thus, we ended up not going with the alternative dataset for this paper, but there are variables in the alternative dataset that may motivate future research on this subject. (EDIT TO MAKE MORE CLEAR)

The paper uses the R programming language [@citeR] to analyze the dataset. The tidyverse package was used to simulate the dataset [@tidyverse]. Also, the tidyverse [@tidyverse], here [@here], arrow [@citearrow] and opendatatoronto [@citeopendatatoronto] packages were used to download the Victims of Crime dataset. Then, the tidyverse [@tidyverse] package was used to clean the raw dataset and generate tests. The testthat package [@citetestthat] was used to create tests for our cleaned dataset. rstanarm [@rstanarm], Arrow [@citearrow], and bayestestR [@citebayestestr] were used to create and test the model. Finally, ggplot2 [@ggplot2], tidyverse [@tidyverse], knitr [@citeknitr] and scales [@citescales] packages were used to create the tables and graphs to display the results. [edit this paragraph]

## Measurement {#sec-data-measurement}

Transforming a real-life Motor Vehicle Collision to an entry in the dataset is a well-documented process by the Toronto Police Service. For insurance purposes, the Toronto Police Service requires drivers to fill out the Motor Vehicle Collision Report for any collisions that occur in Toronto [@tps_mvc_reports]. Drivers required to fill out a motor vehicle collision report if the combined damage is more than $2000, if someone is injured, if a criminal act such as a DUI occurs, or if a pedestrian is involved in the accident [@tps_mvc_reports]. These reports are retained for six years by the Toronto Police Service, with the exception of collisions resulting in a fatality, which are retained indefinitely. The form ensures documentation of collision characteristics, location, road condition, and the extent of damages, systematically recording the characteristics of each crash for further criminal investigation and data analysis.

@@

## Outcome variables {#sec-data-outcome}

@@

```
However, after the transformation, the outcome variable now takes on binary values and the distribution and summary statistics are shown below in @fig-fatalities-cleaned and @tbl-fatalities:

@@

```
## Predictor variables {#sec-data-predictor}

### Hour {#sec-data-predictor-hour}

@@

### Type of Crash {#sec-data-predictor-crash}

@@

### Automobile {#sec-data-predictor-automobile}

@@

### Motorcycle {#sec-data-predictor-motorcycle}

@@

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
### Passenger {#sec-data-predictor-passenger}

@@

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
### Pedestrian {#sec-data-predictor-pedestrian}

@@

# Display the summary statistics in a nicely formatted table
summary_stats %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
### Police Division {#sec-data-predictor-police}

@@

### Year {#sec-data-predictor-year}

@@

### Interaction Terms {#sec-data-predictor-interaction}

@@

#### Hour and Motorcycle {#sec-data-predictor-interaction-hourmotorcycle}

@@

#### Hour and Pedestrian {#sec-data-predictor-interaction-hourpedestrian}

@@

# Display the summary statistics for pedestrians in a formatted table
summary_stats_pedestrian %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```
# Model {#sec-model}

## Model Set-Up and Justification {#sec-model-setupjustification}

@@

## Model Assumptions and Limitations {#sec-model-assumptions}

@@

# Results {#sec-results}

@@

# Discussion {#sec-discussion}

@@

# Appendix {#sec-appendix}

## Exploration of Surveys, Sampling, and Observational Data {#sec-appendix-sso}

### Data Collection Methodology {#sec-appendix-sso-methodology}

The Toronto Police Service's Motor Vehicle Collision database is collected using a systematic process. When a collision occurs such that where persons are injured or a combined damage is valued at more than $\2000 to vehicles or property, or damage to any private, municipal or highway property, police officers complete a standardized Motor Vehicle Accident Report form. This data is then maintained by the City of Toronto, where the collection process follow a standardized process highlighted in the Motor Vehicle Collision Report manual [@mvcr_guide], ensuring reporting consistency. The dataset also defines a collision: the contact resulting from the motion of a motor vehicle or a street car of its load, which produces property damage, injury or death. A standardized definition ensures consistent reporting by the Toronto Police Service and highlights the dataset's clear boundaries.

For every accident that requires one to fill out the Motor Vehicle Accident Report form, the police officer fills out the SR-LD-402 form [@mvcr_guide]. The process ensures consistency in reporting key details surrounding a motor vehicle accident.However, because police officers only need to fill out this form in specific instances, the data set only includes collisions where police officers were present or serious incidents that required a report. As a result, minor offences are usually left out in the dataset and often go unreported.

@@

### Idealized Methodology {#sec-appendix-sso-idealized}

@@

### Example Survey {#sec-appendix-sso-survey}

@@

## Model Card {#sec-appendix-modelcard}

### Model Overview {#sec-appendix-modelcard-overview}
This model predicts the occurrence of motor vehicle collisions (MVCs) based on various factors, including the time of day (hour), type of vehicle involved (motorcycle, pedestrian, automobile, etc.), and police division. The model is trained using data from 2015 to 2024, with a focus on the interactions between the hour of the day and motorcycle or pedestrian involvement.

### Model Details {#sec-appendix-modelcard-details}

@@

### Model Coefficients {#sec-appendix-modelcard-coefficients}

@@

#### Interaction Terms {#sec-appendix-modelcard-interaction}

@@

### Model Performance Metrics {#sec-appendix-modelcard-performance}

@@

### Training Time {#sec-appendix-modelcard-time}
- **Chain 1**:
  - Warm-up: 29.99 seconds
  - Sampling: 26.11 seconds
  - Total: 56.1 seconds
- **Chain 2**:
  - Warm-up: 19.62 seconds
  - Sampling: 17.23 seconds
  - Total: 36.84 seconds
- **Chain 3**:
  - Warm-up: 19.88 seconds
  - Sampling: 19.30 seconds
  - Total: 39.18 seconds
- **Chain 4**:
  - Warm-up: 22.40 seconds
  - Sampling: 19.17 seconds
  - Total: 41.57 seconds

### Conclusion {#sec-appendix-modelcard-conclusion}
The model performs exceptionally well in predicting motor vehicle collisions based on hour, vehicle type, and other factors, with the highest predictive power for motorcycle-related accidents. Key interactions, such as between hour and motorcycle or pedestrian, highlight the importance of specific hours of the day in predicting collision occurrences. This model can be used for more effective traffic safety planning and intervention strategies.

@@

@fig-ppcheck compares observed data (dark line) with replicated posterior predictions (light blue lines). The close alignment suggests that the model accurately captures the data's central tendency and variability. @fig-convergencecheckTrace and @fig-convergencecheckRhat show that the sampling algorithm used, the Markov chain Monte Carlo (MCMC) algorithm, did not run into issues as the posterior distribution for the model was created. Using the checks presented by @tellingstories, both graphs do not show anything abnormal since the trace plots in @fig-convergencecheckTrace display substantial horizontal fluctuation across chains, indicating good mixing, while the Rhat values in @fig-convergencecheckRhat are close to 1 and well below 1.05, further supporting convergence.
